---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Dataset

```{r}
#| label: load-data
#| message: false

matches <- read.csv("data/wcmatches.csv")
wc <- read.csv("data/worldcups.csv")
```

Our data visualization team, RGodz, has elected to use the FIFA World Cup data for our investigation. The 2022 World Cup recently came to a close in Qatar and was full of sensational story lines between FIFA's corruption leading up to the event, the likely end to Cristiano Ronaldo's international football career, and Lionel Messi's fairy-tale victory in the final against a star studded French side led by Kylian Mbappe. Considering all of the sensational headlines and buzz generated by this most recent installment of the World Cup, our team thought this data would be particularly exciting and relevant to analyze.

This FIFA World Cup data was made available on kaggle.com by user Evan Gower and can be found at https://www.kaggle.com/datasets/evangower/fifa-world-cup. The data is broken up into two datasets which we have renamed to `matches` and `wc` . It should be noted that while the most recent World Cup has sparked our interest in the data, the 2022 World Cup data is not included in these datasets and our team will discuss whether or not it makes sense for those observations to be added in the context of our project.

The `matches` dataset includes data from every single game played in the World Cup from the group stage through the final from Uruguay's 1930 tournament to Russia's in 2018. Its variables include specifications such as which teams played in a given match through the variables `home_team` and `away_team` as well as `outcome` which states whether the match ended in a win for the home team, a win for the away team, or a draw. The `matches` dataset has 15 variables (which are all explained in the README.md file of our team's data folder) and thus 15 columns along with 900 rows/observations representing all of the World Cup matches played from 1930 to 2018.

Our second dataset, `wc`, includes more general information about these individual World Cups. It holds a variety of variables such as which nation hosted each tournament with variable `host` , which year it was played with variable `year` , and more information on which nations placed first through fourth. The `wc` dataset has 10 columns for the 10 variables (which are also explained in the README.md file of our team's data folder) and has 21 rows/observations representing the 21 World Cups that were hosted from 1930 to 2018.

# Questions

The two questions you want to answer.

# Analysis plan

A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
